Palomar 5 Gaps

### 2024-10-31 paper/main.tex
Implemented alessandra's comments

  Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and spectroscopy. The development and deployment of the Streamfinder algorithm has led to the detection of about 90 stellar streams in the Milky Way. \citet{2023MNRAS.520.5225M} provides a summary of these streams and tabulates that about 20 are associated to modern day globular clusters. Thus, currently there are about 70 progenitor-less streams and about 145 stream-less clusteres. The solo-streams can be explained by a larger globular cluster population whose memebers have been disolved due to two body relaxation and tidal stripping. For instance, Z Wan 2020( can't find the citation???) reports that the Phenoix stream is dynamically consistent from a globular cluster yet far more metal poor than any cluster ever observed, galactic or otherwise. This implies that the stream is from a completely disolved globular cluster which comes from a population of clusters that have not survived until modern times.

  \citet{2022MNRAS.516.5331M} showed that many very metal poor streams exist in the Milky Way. On the other hand, it is possible that some globular clusters do not loose a significant number of stars to create tidal tails. For instance, \citet{2022A&A...667A.112V} demonstrated that clusters existing in dark matter subhaloes can place the tidal radius far beyond their half mass radii and sheild stars from escape. Another factor is indeed detectability. Additionally, there are mant cluster in the internal regional of the galaxy, a region overwhich \texttt{streamfinder} was not applied since the high stellar density implicates a computational cost that is currently far to expensive. Also, \citet{2018MNRAS.474.2479B} demonstrates that internal dynamics preferentially kicks out small stars first, which have lower luminosities. Earlier escape times places larger distances to the cluster and make detection more difficult.

@@

  Previous studies have predominantly focused on the creation of gaps in stellar streams due to dark matter subhalo passages, with some exceptions that considered giant molecular clouds. If globular clusters can indeed create such gaps, this factor must be taken into account when using stellar streams to infer the population of dark matter subhaloes. In our study, we will fill this gap in the literature by simulating the creation of gaps in the Palomar 5~stream from the galactic globular clusters. We choose Palomar~5 since it is one of the most extensively studied and well-constrained globular cluster streams, known for its ease of observation. We present a result from one of our simulations in Fig. \ref{fig:stream_on_sky}. The following sections detail our modeling of the stream, its gravitational interactions with the galaxy and other globular clusters, and the statistical analysis of perturbations impacting the stream.

@@

    For the galactic potential, we used the second static model from \citet{2017A&A...598A..66P}, a superposition of a thin disk, thick disk, and dark matter halo, with masses and scale lengths provided in Table 1 of \citet{2023A&A...673A..44F}. This model is time-independent throughout our simulations. To balance data volume and computation time, we use only one galactic potential, which should suffice. \citet{2021MNRAS.505.5978V} found that only a few outer globular clusters are strongly affected by different potential models, with kinematic uncertainties being the dominant factor in orbital solution differences. Similarly, \citet{2024MNRAS.528.5189G} generated a globular cluster mass-loss catalog using seven different potential models and found their debris distributions to be model-independent. While the clusters' positions may depend on the model, we assert that interaction rates and stream formation are unaffected by the choice of galactic model.

    The primary methodological departure from \citet{2023A&A...673A..44F} is that now we include globular cluster interactions. Instead of treating the clusters as point masses that only experience the gravitational potential of the galaxy, we now compute N-body interactions between the clusters during the backward integration over 5 billion years. The equations of motion for the globular cluster system are thus: 
    \begin{equation}
      \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
    \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galacto-centric position, the index $i$ indicates a globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates. The index $j$ indicates the other globular cluster that are summed over where: $m_j$ is its mass, $b_j$ is its Plummer scale radius, and $r_j$ is it's galacto-centric position. $\Phi$ is the axis-symmteric potential of the model used for the Milky Way.

    At their computed positions 5 Gyr ago, Palomar 5 was assigned 100,000 particles sampled from a Plummer distribution to represent its stellar content. For each set of initial conditions, the particle were resampled based on uncertainties in mass, which is an improvement from \citet{2023A&A...673A..44F}.

    We use a mass estimate for Palomar 5 taken from \citet{2021MNRAS.505.5957B}, $1.34\pm0.24\times10^{4}M_{\odot}$--which is a contested value.

@@

    We then integrate the evolution of these particles forward in time to the present day. During this forward integration, each star particle feels the gravitational influence of the galaxy and all globular clusters, but not other star particles. We do not perform direct N-body simulations between star particles as in eq.~\ref{eq:GCNBody}; in other words, we are using test-particle and solving the restricted three body problem. At each time step, we load the positions of all globular clusters, as found from the initial integration, and apply their forces to the star particles. The equation of motion is thus: 
    \begin{equation}
      \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right),
      \end{equation} \label{eq:equation_of_motion_particle} where the index $p$ represents one of the 100,000 particles of interest.

@@

    The mass-size relation, shown in the top-right panel, reveals a lower limit of $10^5~\textrm{M}_{\odot}$ for clusters that provoke gaps. While mass and size alone are insufficient predictors, this empirical limit is useful because relative velocity and impact parameters are historical values. The bottom-left panel offers a more physically meaningful result: gap-creating impacts tend to impart more momentum into the stream than non-gap impacts, though with some overlap. While \citet{2015MNRAS.450.1136E} extended the three-parameter model (Eq.~\ref{eq:change_in_momentum}) to include the size of the perturber and the perpendicular and parallel components of relative velocity, we found no clear combination of these factors that distinguishes gap from non-gap cases.

    Further insights are shown in Fig.\ref{fig:impact_geometry_statistics}, highlighting a maximum impact parameter of around 200~pc, roughly the width of the stream. Additionally, as \citet{2016MNRAS.457.3817S} demonstrated, the dispersion of action-frequencies in the stream plays a role. Regions of the stream with well-separated frequencies are more susceptible to gap formation, while those with a wide frequency range (near the cluster) tend to erase the history of impacts. Thus, for a gap to form, the imparted change in frequency must exceed the range of frequencies in the impacted region. Thus, the strong flybys that occured close to the cluster and were thus inconsequential for gap formation.

@@

  \begin{itemize}
    \item \textbf{Orbital Overlap}: clusters with orbital energies and angular momenta that overlap with Palomar 5's orbit can impact its stream.
    \item \textbf{Mass Threshold}: Perturbing clusters have masses above $10^5 M_\odot$; no perturber has a mass smaller than this.
    \item \textbf{Galactic radius}: Impacts preferentially occur at shorter galactic radii.
    \item \textbf{Momentum and Frequency Change}: Gap creation is linked to the momentum imparted to the stream during the flyby, yet not entirely. As showcased by \citet{2016MNRAS.457.3817S}, the change in orbital frequencies induced by the flyby must be greater than the spread in frequencies present in the stream. Some impacts imparted significant momentum changes that did not lead to gap growth for this reason.
  \end{itemize}

@@

  On the other hand, for a stream length of 10~kpc reported from \citet{2020ApJ...889...70B} and given our gap creation rate, in 5~Gyr we would expect to see about 0.5 gaps, in 10~Gyr's we would expect 1. Therefore, finding two gaps from globular clusters would not be impossible.

>>>>

  Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and spectroscopy. The development and deployment of the Streamfinder algorithm has led to the detection of about 90 stellar streams in the Milky Way. \citet{2023MNRAS.520.5225M} provides a summary of these streams and tabulates that about 20 are associated to modern day globular clusters. Thus, currently there are about 70 progenitor-less streams and about 145 stream-less clusteres. The solo-streams can be explained by a larger globular cluster population whose members have been disolved due to two body relaxation and tidal stripping. For instance, Z Wan 2020( can't find the citation???) reports that the Phenoix stream is dynamically consistent from a globular cluster yet far more metal poor than any cluster ever observed, galactic or otherwise. This implies that the stream is from a completely disolved globular cluster which comes from a population of clusters that have not survived until modern times.

  \citet{2022MNRAS.516.5331M} showed that many very metal poor streams exist in the Milky Way. On the other hand, it is possible that some globular clusters do not loose a significant number of stars to create tidal tails. For instance, \citet{2022A&A...667A.112V} demonstrated that clusters existing in dark matter subhaloes can place the tidal radius far beyond their half mass radii and sheild stars from escape. Another factor is indeed detectability. Additionally, there are many cluster in the internal regional of the galaxy, a region overwhich \texttt{streamfinder} was not applied since the high stellar density implicates a computational cost that is currently far to expensive. Also, \citet{2018MNRAS.474.2479B} demonstrates that internal dynamics preferentially kicks out small stars first, which have lower luminosities. Earlier escape times places larger distances to the cluster and make detection more difficult.

@@

  Previous studies have predominantly focused on the creation of gaps in stellar streams due to dark matter subhalo passages, with some exceptions that considered giant molecular clouds. To our knowledge, no study has yet been perfomed investigating the possibility of globular clusters greating gaps in streams. If globular clusters can indeed create such gaps, this factor must be taken into account when using stellar streams to infer the population of dark matter subhaloes. In our study, we will fill this gap in the literature by simulating the creation of gaps in the Palomar 5~stream from the galactic globular clusters. We choose Palomar~5 since it is one of the most extensively studied and well-constrained globular cluster streams, known for its ease of observation. We present a result from one of our simulations in Fig. \ref{fig:stream_on_sky}. The following sections detail our modeling of the stream, its gravitational interactions with the galaxy and other globular clusters, and the statistical analysis of perturbations impacting the stream.

  \salvatore{Alessandra dice che va messo molto di più in evidenza il fatto che nessuno sa se le predizioni cosmologiche sono validi sulla scala galactica riguarda ai sotto aloni. . Forse va cambiato l'ordine del discorso. To give more focus to the intro, I would start introducing GCs, their streams and their use as DM probes. Then you can say something about dark matter, the fact that we don't it's distribution (which makes streas so important) and that we don't know if there are the predicted subhalos. You can continue saying that gap in GCs have been used to put constraints on these subhalo number but... and then you continue with the literature, what is missing and what you did.}

@@

    For the galactic potential, we used the second static model from \citet{2017A&A...598A..66P}, a superposition of a thin disk, thick disk, and dark matter halo, with masses and scale lengths provided in Table 1 of \citet{2023A&A...673A..44F}. This model is time-independent throughout our simulations. \salvatore{This model satisfies a series of observational constraints such as local solar stellar density, the galactic rotation curve, etc, similarly to other galactic models such as Bovy2014, McMillian.} However, To balance data volume and computation time, we use only one galactic potential, which should suffice. \citet{2021MNRAS.505.5978V} found that only a few outer globular clusters are strongly affected by different potential models, with kinematic uncertainties being the dominant factor in orbital solution differences. Similarly, \citet{2024MNRAS.528.5189G} generated a globular cluster mass-loss catalog using seven different potential models and found their debris distributions to be model-independent, similarly to \citet{2023A&A...673A..44F}. While the clusters' positions may depend on the model, we assert that interaction rates and stream formation are unaffected by the choice of galactic model.

    There are two primary methodological departures from \citet{2023A&A...673A..44F}. Previously, the clusters only experienced the galaxy. Now, we include N-body forces between the clusters. Secondly, the clusters are not point masses but are softened as Plummer spheres each with their own scale radii. The equations of motion for the globular cluster system are thus: 
    \begin{equation}
      \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
    \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galacto-centric position, the index $i$ indicates a globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates. The index $j$ indicates the other globular cluster that are summed over where: $m_j$ is its mass, $b_j$ is its Plummer scale radius, and $r_j$ is it's galacto-centric position. $\Phi$ is the axis-symmteric potential of the model used for the Milky Way. Note that the mass and scale radii are constant in these simulations and we do not take into account the internal evolution of the globular clusters.

    At the end of this integration we have positions and velocities of the entire globular cluster system 5 Gyr ago. We sample 100,000 particles from a Plummer distribution using the same total mass and characteristic radius for the orbital integration. We repeate this sampling for each of the 50 Monte-Carlo initial conditions. Considering the mass uncertainties is an improvement from \citet{2023A&A...673A..44F} where the most probable mass was used for each Monte-Carlo sampling.

    Regarding Palomar 5's mass, we use a mass estimate for Palomar 5 taken from \citet{2021MNRAS.505.5957B}, $1.34\pm0.24\times10^{4}M_{\odot}$--which is a contested value.

@@

    We then integrate the evolution of these particles forward in time to the present day. During this forward integration, each star particle feels the gravitational influence of the galaxy and all globular clusters, but not other star particles. We do not perform direct N-body simulations between star particles as we did for globular clusters. In other words, we are using test-particle and solving the restricted three body problem. The particles do not exert/feel gravitational forces upon/from one another. At each time step, we load the positions of all globular clusters, as found from the initial integration, and apply their forces to the star particles. The equation of motion is thus: 
    \begin{equation}
      \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right),
      \end{equation} \label{eq:equation_of_motion_particle} where the index $p$ represents one of the 100,000 particles of interest, and $j$ indexes over the globular clusters as in Eq.~\ref{eq:GCNBody}

@@

    The mass-size relation, shown in the top-right panel, reveals a lower limit of $10^5~\textrm{M}_{\odot}$ for clusters that provoke gaps. While mass and size alone are insufficient predictors, this empirical limit is useful because relative velocity and impact parameters are inacessible. Concentration or the density profile of the perturber has a must less significant effect respect to the total mass and subsequent evolution of the gap as shown in Fig.~3 of \citet{2016MNRAS.457.3817S}. The bottom-left panel offers a more physically meaningful result: gap-creating impacts tend to impart more momentum into the stream than non-gap impacts, though with some overlap. While \citet{2015MNRAS.450.1136E} extended the three-parameter model (Eq.~\ref{eq:change_in_momentum}) to include the size of the perturber and the perpendicular and parallel components of relative velocity, we found no clear combination of these factors that distinguishes gap from non-gap cases.

    Further insights are shown in Fig.\ref{fig:impact_geometry_statistics}, highlighting a maximum impact parameter of around 200~pc, roughly the width of the stream. Additionally, as \citet{2016MNRAS.457.3817S} demonstrated, the dispersion of action-frequencies in the stream plays a role. The frequencies being the rate at which a particle completes a cycle in phase space. For instance, the frequency corresponding to the azimuthal action, $J_\phi$, is $\dot\theta_\phi = - \frac{\mathcal{\partial H}}{\partial J_\phi}$ and where $\theta_\phi$ is the angle describing the position of the particle phase space between momentum ($p_\phi$) and position ($\phi$) axes---where $\phi$ is the azimuhtal angle between the x-y axes in physical space. Regions of the stream with well-separated frequencies are more susceptible to gap formation, while those with a wide frequency range (near the cluster) tend to erase the history of impacts. Thus, for a gap to form, the imparted change in frequency must exceed the range of frequencies in the impacted region. Thus, the strong flybys that occured close to the cluster and were thus inconsequential for gap formation.

@@

  \begin{itemize}
    \item \textbf{Orbital Overlap}: clusters with orbital energies and angular momenta that overlap with Palomar 5's orbit can impact its stream.
    \item \textbf{Mass Threshold}: Perturbing clusters have masses above $10^5 M_\odot$; no perturber has a mass smaller than this.
    \item \textbf{Galactic radius}: Impact occurrences negatively correlates with galacto-centric distance.
    \item \textbf{Momentum and Frequency Change}: Gap creation is linked to the momentum imparted to the stream during the flyby, yet not entirely. As showcased by \citet{2016MNRAS.457.3817S}, the change in orbital frequencies induced by the flyby must be greater than the spread in frequencies present in the stream. Some impacts imparted significant momentum changes that did not lead to gap growth for this reason.
  \end{itemize}

@@

  On the other hand, to provide a crude estiamte, consider  a stream length of 10~kpc reported from \citet{2020ApJ...889...70B} and given our gap creation rate. In 5~Gyr we would expect to see about 0.5 gaps, in 10~Gyr's we would expect 1. Therefore, finding two gaps from globular clusters would not be impossible.
